---
title: "Automatically linking output files to a Quarto document during render"
vignette: >
  %\VignetteIndexEntry{Automatically linking output files to a Quarto document during render}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
theme: slate
knitr:
  opts_chunk:
    collapse: true
    results: hold
---

```{r}
#| label: setup
library(promicrobialTools)
```

```{r}
#| eval: false
#| echo: false
#| results: asis
# Function to save data and create link
save_and_link <- function(data, filename, description = NULL) {
  # Get current file directory
  current_file <- knitr::current_input()
  if (!is.null(current_file)) {
    file_dir <- dirname(current_file)
  } else {
    file_dir <- getwd()
  }

  # Create results directory
  results_dir <- file.path(file_dir, "results")
  if (!dir.exists(results_dir)) {
    dir.create(results_dir, recursive = TRUE)
  }

  # Full file path
  file_path <- file.path(results_dir, filename)

  # Save the data based on file extension
  ext <- tools::file_ext(filename)

  if (ext == "csv") {
    write.csv(data, file_path, row.names = FALSE)
  } else if (ext == "xlsx") {
    openxlsx::write.xlsx(data, file_path)
  } else if (ext == "rds") {
    saveRDS(data, file_path)
  } else if (ext == "txt") {
    writeLines(as.character(data), file_path)
  }

  # Generate markdown link
  relative_path <- file.path("results", filename)
  link_text <- if (!is.null(description)) {
    description
  } else {
    paste("Download", filename)
  }

  # Return markdown link
  cat(paste0("[", link_text, "](", relative_path, ")\n\n"))

  return(invisible(file_path))
}

generate_results_section <- function(results_dir = "results") {
  current_file <- knitr::current_input()
  if (!is.null(current_file)) {
    base_dir <- dirname(current_file)
  } else {
    base_dir <- getwd()
  }

  full_results_dir <- file.path(base_dir, results_dir)

  if (!dir.exists(full_results_dir)) {
    return(invisible(NULL))
  }

  files <- list.files(full_results_dir, full.names = FALSE)

  if (length(files) == 0) {
    return(invisible(NULL))
  }

  cat("\n## 📊 Analysis Results\n\n")
  cat("The following files contain detailed results from this analysis:\n\n")

  # Group files by type
  file_types <- list(
    "Data Files" = c("csv", "xlsx", "rds", "json"),
    "Plots" = c("png", "pdf", "svg", "jpg", "jpeg"),
    "Reports" = c("html", "docx", "txt", "md"),
    "Other" = character(0)
  )

  files_by_type <- list()

  for (file in files) {
    ext <- tools::file_ext(tolower(file))
    assigned <- FALSE

    for (type_name in names(file_types)) {
      if (ext %in% file_types[[type_name]]) {
        if (is.null(files_by_type[[type_name]])) {
          files_by_type[[type_name]] <- character(0)
        }
        files_by_type[[type_name]] <- c(files_by_type[[type_name]], file)
        assigned <- TRUE
        break
      }
    }

    if (!assigned) {
      if (is.null(files_by_type[["Other"]])) {
        files_by_type[["Other"]] <- character(0)
      }
      files_by_type[["Other"]] <- c(files_by_type[["Other"]], file)
    }
  }

  # Generate links by type
  for (type_name in names(files_by_type)) {
    if (length(files_by_type[[type_name]]) > 0) {
      cat(paste0("\n### ", type_name, "\n\n"))

      for (file in files_by_type[[type_name]]) {
        file_path <- file.path(results_dir, file)
        full_file_path <- file.path(full_results_dir, file)

        # Get file size
        if (file.exists(full_file_path)) {
          size_kb <- round(file.size(full_file_path) / 1024, 1)
          size_info <- paste0(" (", size_kb, " KB)")
        } else {
          size_info <- ""
        }

        # Generate appropriate icon
        ext <- tools::file_ext(tolower(file))
        icon <- switch(
          ext,
          "csv" = "📊",
          "xlsx" = "📋",
          "rds" = "💾",
          "png" = "🖼️",
          "pdf" = "📄",
          "svg" = "🎨",
          "html" = "🌐",
          "txt" = "📝",
          "md" = "📄",
          "📄"
        )

        cat(paste0(
          "- ",
          icon,
          " [",
          tools::file_path_sans_ext(file),
          "](",
          file_path,
          ")",
          size_info,
          "\n"
        ))
      }
    }
  }

  cat("\n")
}

```

## Analysis

First lets create some simple results using the `mtcars` dataset from the [Tidyverse package](https://www.tidyverse.org/).

```{r analysis}
#| results: asis

# Perform analysis
result_data <- mtcars %>%
  group_by(cyl) %>%
  summarise(
    mean_mpg = mean(mpg),
    mean_hp = mean(hp),
    count = n()
  )

# Show first few rows
kable(head(result_data))
```

Using the `save_and_link` function, we can automatically save the results to a folder called "results" in the project root directory, with either the name of the current file's parent directory or the name of the file itself. For example, 

:::{.column-page}
::: {.grid}

::: {.g-col}

If you chose to store results on a parent folder level, say for groupinig particular parts of your analysis, you would end up with something like this:

```
project_root/
├── _quarto.yml
├── analysis/
│   └── data-quality-control/
│       └── data-filtering.qmd
│       └── metadata-cleaning.qmd
│   └── regression/
│       └── zero-inflated-model.qmd   # Current file
└── results/
    └── regression/     # Based on parent folder
        ├── analysis_results.csv
        ├── sample_data.xlsx
        └── summary_stats.txt
```

:::

::: {.g-col}

If you store results on a file level (using `use_filename = TRUE`), then you would end up with a directory structure like this:

```
project_root/
├── _quarto.yml
├── analysis/
│   └── data-quality-control/
│       └── data-filtering.qmd
│       └── metadata-cleaning.qmd
│   └── regression/
│       └── zero-inflated-model.qmd   # Current file
└── results/
    └── zero-inflated-model/     # Based on file name
        ├── analysis_results.csv
        ├── sample_data.xlsx
        └── summary_stats.txt
```
:::

:::
:::

For the sake of this vignette, results are going to be saved in a folder named after the parent folder (the default), which will be "vignettes". The `save_and_link` function can handle most data types.

```{r}
#| results: asis
# Save full results and create link
save_and_link(
  data = result_data,
  filename = "cylinder_summary.csv",
  description = "🔧 MT Cars Cylinder Analysis Summary",
  use_filename = FALSE
)

save_and_link(
  data = mtcars,
  filename = "raw_data.xlsx",
  description = "📊 Raw Motor Trend Data",
  use_filename = FALSE
)
```

If you were to generate results files outside of the `save_and_link` function, it will still appear in the results section generated by `generate_results_section()`, so long as it is in the same folder. This may be useful for example when you save a plot but don't want to link it because it will be rendered during knit.

You can easily point towards the results folder used by `save_and_link` or `generate_results_section` using the function `get_results_dir`.

```{r}
#| eval: true

filepath <- get_results_dir(use_filename = FALSE)

```

Then you can save plots or other results files without having to worry about making a mistake specifying the right directory for the current file.

```{r}
#| eval: true
#| label: fig-mtcars
#| fig-cap: MPG vs Horsepower
(p <- ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  theme_minimal())
```

```{r}
#| eval: true
#| results: false
png(file.path(filepath, "mtcars-plot.png"), width = 1200, height = 800)
plot(p)
dev.off()
```

When all is said and done, you can summarise the results files generated during the document's render. These will include any saved to the document's folder that weren't saved with `save_and_link`.

```{r}
#| results: asis
# Auto-generate results section at the end
generate_results_section(use_filename = FALSE)
```
