---
title: "Automatically linking output files to a Quarto document during render"
vignette: >
  %\VignetteIndexEntry{Automatically linking output files to a Quarto document during render}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
theme: slate
knitr:
  opts_chunk:
    collapse: true
    results: hold
    comment: ""
---

```{r}
#| label: setup
library(promicrobialTools)
```

## Introduction

This vignette demonstrates an enhanced file organization system that provides three flexible options for organizing analysis results. The system automatically creates structured directories and generates download links for files created during analysis.

## Directory Organization Options

The enhanced system offers three directory naming strategies:

:::{.column-page}
::: {.grid}

::: {.g-col-4}
### Option 1: Parent Directory (`"parent"`)
**Default behavior** - Groups results by analysis section:

```
project_root/
├── _quarto.yml
├── analysis/
│   └── regression/
│       └── linear-model.qmd   # Current file
└── results/
    └── regression/     # Based on parent folder
        ├── model_results.csv
        ├── diagnostics.png
        └── summary.rds
```

**Best for:** Organizing by analysis type or project section
:::

::: {.g-col-4}
### Option 2: Filename (`"filename"`)
**File-specific organization** - Each document gets its own folder:

```
project_root/
├── _quarto.yml
├── analysis/
│   └── regression/
│       └── linear-model.qmd   # Current file
└── results/
    └── linear-model/     # Based on filename
        ├── model_results.csv
        ├── diagnostics.png
        └── summary.rds
```

**Best for:** Independent analyses or avoiding file conflicts
:::

::: {.g-col-4}
### Option 3: Nested Structure (`"both"`)
**Maximum organization** - Nested parent/filename structure:

```
project_root/
├── _quarto.yml
├── analysis/
│   └── regression/
│       └── linear-model.qmd   # Current file
└── results/
    └── regression/         # Parent directory
        └── linear-model/   # Filename subdirectory  
            ├── model_results.csv
            ├── diagnostics.png
            └── summary.rds
```

**Best for:** Complex projects maintaining both context and specificity
:::

:::
:::

## Example Analysis

Let's demonstrate all three approaches using the `mtcars` dataset from the [Tidyverse package](https://www.tidyverse.org/).

```{r analysis}
#| results: asis

# Perform analysis
result_data <- mtcars %>%
  group_by(cyl) %>%
  summarise(
    mean_mpg = mean(mpg),
    mean_hp = mean(hp),
    mean_wt = mean(wt),
    count = n(),
    .groups = 'drop'
  )

# Show results
kable(result_data, caption = "Summary statistics by cylinder count")
```

Using the `save_and_link` function, we can automatically save the results to a folder called "results" in the project root directory, with either the name of the current file's parent directory, the name of the file itself, or both. 

### Method 1: Parent Directory Organization

Using the default `"parent"` option groups results by the parent directory name:

```{r}
#| results: asis

# Save using parent directory naming (default)
save_and_link(
  data = result_data,
  filename = "cylinder_summary_parent.csv",
  description = "🔧 Cylinder Summary (Parent Directory Method)",
  directory_naming = "parent" # This is the default
)

# You can also omit the parameter for default behavior
save_and_link(
  data = mtcars[1:10, ],
  filename = "sample_data_parent.xlsx",
  description = "📊 Sample Data (Parent Directory Method)"
)
```

### Method 2: Filename-Based Organization

Using `"filename"` creates a directory named after the current document:

```{r}
#| results: asis

save_and_link(
  data = result_data,
  filename = "cylinder_summary_filename.csv",
  description = "🔧 Cylinder Summary (Filename Method)",
  directory_naming = "filename"
)

save_and_link(
  data = mtcars[11:20, ],
  filename = "sample_data_filename.xlsx",
  description = "📊 Sample Data (Filename Method)",
  directory_naming = "filename"
)
```

### Method 3: Combined Naming

Using `"both"` creates the most specific directory structure with a folder named after the current file's parent folder and a subdirectory using the file's name:

```{r}
#| results: asis

save_and_link(
  data = result_data,
  filename = "cylinder_summary_combined.csv",
  description = "🔧 Cylinder Summary (Combined Method)",
  directory_naming = "both"
)

save_and_link(
  data = mtcars[21:30, ],
  filename = "sample_data_combined.xlsx",
  description = "📊 Sample Data (Combined Method)",
  directory_naming = "both"
)
```

## Working with Different Directory Paths

If you save files outside of the `save_and_link` function but want to use the same folder structure, you can easily point towards the results folder using the function `get_results_dir`.

```{r}
#| echo: true

# Show the different paths that would be created
cat("Parent directory method:", get_results_dir("parent"), "\n")
cat("Filename method:", get_results_dir("filename"), "\n")
cat("Combined method:", get_results_dir("both"), "\n")
```

Now you can save plots or other results files without having to worry about making a mistake specifying the right directory for the current file.

```{r}
#| label: fig-mpg-hp
#| fig-cap: "MPG vs Horsepower by Cylinder Count"

# Create plot
mpg_plot <- ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Fuel Efficiency vs Engine Power",
    x = "Horsepower",
    y = "Miles per Gallon",
    color = "Cylinders"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(mpg_plot)
```

Now let's save this plot using each directory method:

```{r}
#| results: false

# Save plots to different directories
ggsave(
  filename = file.path(get_results_dir("parent"), "mpg_vs_hp_parent.png"),
  plot = mpg_plot,
  width = 10,
  height = 6,
  dpi = 300
)

ggsave(
  filename = file.path(get_results_dir("filename"), "mpg_vs_hp_filename.png"),
  plot = mpg_plot,
  width = 10,
  height = 6,
  dpi = 300
)

ggsave(
  filename = file.path(get_results_dir("both"), "mpg_vs_hp_combined.png"),
  plot = mpg_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

## Multiple file types are supported

The file type is automatically detected by the extension in the filename provided in the function, and saved with the appropraite function.

Below is an example showing saving an R object as an `.rds` file

```{r}
#| results: asis

# Fit different models
linear_model <- lm(mpg ~ hp + wt + factor(cyl), data = mtcars)
simple_model <- lm(mpg ~ hp, data = mtcars)

# Save models using different directory structures
save_and_link(
  data = linear_model,
  filename = "linear_model.rds",
  description = "📈 Multi-variable Linear Model",
  directory_naming = "parent"
)
```

## Copying External Files

You can also copy existing files to results directories:

```{r}
#| eval: false
#| results: asis

# Example of copying external files (not run in vignette)
external_files <- c("data/external_dataset.csv", "docs/methodology.pdf")
descriptions <- c("📊 External Dataset", "📄 Analysis Methodology")

# Copy to parent directory structure
copy_to_results(
  source_files = external_files,
  directory_naming = "parent",
  descriptions = descriptions
)

# Copy to combined directory structure
copy_to_results(
  source_files = external_files,
  directory_naming = "both",
  descriptions = descriptions
)
```

## Cleaning Up Results

Each directory method has its own cleanup function:

```{r}
#| eval: false

# Clean specific directory types (not run in vignette)
clean_results_dir("parent", confirm = TRUE)
clean_results_dir("filename", confirm = TRUE)
clean_results_dir("both", confirm = TRUE)
```

## Summary: Generated Results by Method

When all is said and done, you can summarise the results files generated during the document's render. These will include anything saved in the document's folder regardless of whether they were saved with `save_and_link`.

### Parent Directory Results
```{r}
#| results: asis
generate_results_section("parent")
```

### Filename-Based Results  
```{r}
#| results: asis
generate_results_section("filename")
```

### Combined Directory Results
```{r}
#| results: asis
generate_results_section("both")
```

## Overview of All Results

Finally, let's see all results directories that were created:

```{r}
#| results: asis
list_all_results()
```

## Best Practices and Recommendations

### When to Use Each Method:

**Parent Directory (`"parent"`)**:
- ✅ Default choice for most analyses
- ✅ When organizing by analysis type (e.g., "preprocessing", "modeling", "validation")
- ✅ Team projects where results should be grouped by workflow stage
- ✅ When multiple documents in the same folder should share results

**Filename (`"filename"`)**:
- ✅ Independent analyses that shouldn't be mixed
- ✅ When each document represents a complete, standalone analysis
- ✅ Avoiding filename conflicts in collaborative projects
- ✅ When you want clear 1:1 mapping between documents and results

**Nested Structure (`"both"`)**:
- ✅ Large, complex projects with many similar analyses
- ✅ When you need maximum organization specificity
- ✅ Longitudinal studies with repeated analyses
- ✅ When both context (parent) and specificity (filename) are important

This enhanced system provides the flexibility to organize results according to your project's specific needs while maintaining clean, predictable directory structures.