---
title: "Cleaning your data before starting analysis"
vignette: >
  %\VignetteIndexEntry{Cleaning your data before starting analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
author: "Nathaniel Cole"
date: "`r Sys.Date()`"
theme: slate
knitr:
  opts_chunk:
    collapse: true
    results: hold
    comment: ""
---

```{r}
#| label: setup
library(promicrobialTools)
```

## Text Case Manipulation with capwords

The `capwords` function provides flexible text case manipulation, particularly useful for standardizing titles, names, and other text data. Let's explore its various features through some practical examples.

```{r}
# Basic Usage
text <- c("the quick brown fox", "R programming LANGUAGE", "machine LEARNING")
capwords(text)
```

### Strict Mode

When `strict = TRUE`, all characters after the first letter are converted to lowercase

```{r}
capwords(text, strict = TRUE)

```

### Working with Different Separators
```{r}
hyphenated <- c("self-driving", "state-of-the-art", "real-time-processing")
capwords(hyphenated, split = "-")
```

### Preserving Special Terms
```{r}
hyphenated <- c("self-driving", "state-of-the-art", "real-time-processing")
capwords(hyphenated, split = "-")
```

### Handling Mixed Input Types

```{r}
mixed_input <- c("john doe", NA, "JANE smith", factor("bob wilson"))
capwords(mixed_input, strict = TRUE)
```

### Using with dplyr

```{r}
if (require(dplyr)) {
  authors %>%
    mutate(
      name = capwords(name, strict = TRUE),
      institution = capwords(institution)
    )
}
