---
title: "Creating Phyloseq Summary Tables"
vignette: >
  %\VignetteIndexEntry{Creating Phyloseq Summary Tables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
author: "Nathaniel Cole"
date: "`r Sys.Date()`"
theme: slate
knitr:
  opts_chunk:
    collapse: true
    results: hold
    comment: ""
---

## Introduction

The `phyloseq_summary()` function provides a comprehensive summary of phyloseq objects that is essential for reproducible microbiome data analysis. This function generates professional-looking summary tables that are compatible with both HTML and PDF rendering in Quarto documents.

```{r}
#| label: setup
library(promicrobialTools)
library(phyloseq)
```

## Loading Example Data

We'll use the built-in GlobalPatterns dataset from phyloseq for demonstration:

```{r}
# Load example data
data("GlobalPatterns")
gp <- GlobalPatterns

# Let's also create a subset for faster processing
gp_subset <- subset_samples(
  GlobalPatterns,
  SampleType %in% c("Feces", "Skin", "Tongue")
)
gp_subset <- prune_taxa(taxa_sums(gp_subset) > 100, gp_subset)
```

## Default GT Format (Best for HTML)

The default format creates a styled gt table that renders beautifully in HTML documents:

```{r}
# Create summary table with default settings
phyloseq_summary(gp_subset)
```

## Kable Format (Best for PDF)

For PDF rendering, use the kable format which provides better compatibility:

```{r}
# Create summary table for PDF rendering
summary_kable <- phyloseq_summary(
  gp_subset,
  format = "kable",
  title = "GlobalPatterns Dataset Summary"
)

summary_kable
```

## Data Frame Format (For Further Processing)

Get the raw data frame for custom processing:

```{r}
# Get summary as data frame
summary_df <- phyloseq_summary(gp_subset, format = "df")
print(summary_df)
```

# Understanding the Output

The function provides three main sections of information:

## 1. Dataset Structure (📊)

This section provides essential metadata about the phyloseq object:

- **Number of samples**: Total sample count (NA if not available)
- **Number of taxa**: Total taxa/OTU count (NA if not available)  
- **Number of taxonomic ranks**: Count of taxonomic levels
- **Taxonomic ranks**: List of available taxonomic ranks (e.g., Kingdom, Phylum, etc.)
- **Phylogenetic tree**: Whether a phylogenetic tree is present (Present/Absent)
- **Abundance type**: Whether abundances are Relative, Absolute, or Unknown

## 2. Read Statistics (📈)

Standard sequencing depth and distribution metrics:

- Read count ranges (min/max per sample)
- Total, average, and median read counts
- Essential for understanding sequencing effort and depth

## 3. Data Quality Metrics (🔍)

Quality assessment metrics critical for downstream analysis:

- **Data sparsity**: Percentage of zero values in the OTU table
- **Singletons**: OTUs with only one read across all samples
- Important for filtering decisions and analysis interpretation

# Advanced Usage

## Custom Titles and Formatting

```{r eval=FALSE}
# Custom title and subtitle
phyloseq_summary(
  gp_subset,
  title = "Human Microbiome Analysis",
  subtitle = "16S rRNA gene sequencing results from multiple body sites",
  decimal_places = 2
)
```

## Best Practices

### **Quality Assessment**
- Pay attention to sparsity levels (>80% may indicate over-filtering)
- Check singleton percentages (high values may suggest contamination)
- Verify abundance types match your expectations

### **Reproducibility**
- Include phyloseq object summaries at multiple analysis stages
- Document any filtering or transformation steps
- Use consistent decimal places for comparability

## Troubleshooting

### Common Issues

1. **"Package not installed" errors**: Ensure all dependencies are installed
2. **GT tables not rendering in PDF**: Use `format = "kable"` instead
3. **"NA" values in output**: Normal for incomplete phyloseq objects
4. **Abundance type "Unknown"**: May occur with unusual data distributions

### Getting Help

```{r eval=FALSE}
# View function documentation
?phyloseq_summary

# Check phyloseq object structure
str(your_phyloseq_object)

# Validate phyloseq object
phyloseq::validate_phyloseq(your_phyloseq_object)
```

---

This vignette demonstrates the functionality of `phyloseq_summary()` for extracting important dataset metadata from phyloseq objects in microbiome analysis workflows.
