---
title: "Data Wrangling with process_data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Wrangling with process_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(promicrobialTools)
```

## Introduction

The `process_data` function is designed to transform wide-format rating data into a long format, making it easier to analyze and visualize paired ratings from multiple reviewers. This vignette demonstrates how to use the function and shows common use cases.

## Basic Usage

First, let's create some example data:

```{r setup}
library(dplyr)
library(tidyr)

# Create example data
example_data <- data.frame(
  x1 = c(3.0, 2.5, 4.0),
  y1 = c(2.8, 2.3, 3.8),
  x2 = c(2.9, 2.4, 3.9),
  y2 = c(3.1, 2.6, 4.1),
  group = c("A", "A", "B")
)

# View the original data
example_data
```

### Without Grouping

The basic usage transforms the data without preserving any grouping:

```{r}
result1 <- process_data(example_data)
result1
```

### With Grouping

To preserve a grouping variable:

```{r}
result2 <- process_data(example_data, group)
result2
```

## Common Use Cases

### Calculating Agreement Metrics

The long format makes it easy to calculate agreement between reviewers:

```{r}
result2 %>%
  group_by(group) %>%
  summarise(
    mean_diff = mean(abs(x - y)),
    correlation = cor(x, y)
  )
```

### Visualization

The long format is ideal for creating visualizations:

```{r}
library(ggplot2)

ggplot(result2, aes(x = x, y = y, color = group)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  facet_wrap(~reviewer) +
  theme_minimal() +
  labs(title = "Ratings by Reviewer and Group", x = "X Rating", y = "Y Rating")
```

## Error Handling

The function includes several checks to ensure data validity:

```{r, error=TRUE}
# Missing columns
bad_data <- data.frame(x1 = 1:3, y2 = 4:6)
process_data(bad_data)

# Empty data frame
process_data(data.frame())

# Invalid column names
bad_names <- data.frame(xa = 1:3, ya = 4:6)
process_data(bad_names)
```

## Tips and Best Practices

1. Ensure your column names follow the pattern 'x1', 'y1', 'x2', 'y2', etc.
2. Check that all x-y pairs are complete before processing
3. Consider the scale and units of your ratings when interpreting results
4. Use appropriate visualization techniques for your specific type of rating data

## Conclusion

The `process_data` function provides a robust way to transform rating data while maintaining data integrity and offering flexibility in handling grouped data. The resulting long format is ideal for various types of analyses and visualizations.
