---
title: "Scaling and Backtransforming Numeric Data with scale_values"
theme: slate
knitr:
  opts_chunk:
    eval: true
    collapse: true
    results: hold
    comment: ""
author: "Nathaniel Cole"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Scaling and Backtransforming Numeric Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
# Load required libraries
library(promicrobialTools)
library(ggplot2)
library(knitr)

# Set seed for reproducibility
set.seed(123)
```

# Introduction

This vignette demonstrates how to use the scaling functions for normalizing numeric data to the [0,1] range and performing the inverse transformation. The package provides three main functions:

- `scale_values()`: Scale data to [0,1] range
- `get_scaled_value()`: Scale new values using existing scaling parameters  
- `backtransform_scaled_values()`: Convert scaled values back to original scale

These functions are particularly useful for data normalization.

# Basic Usage

## Scaling Simple Vectors

Let's start with basic vector scaling:

```{r basic-scaling}
# Create sample data
original_data <- c(10, 25, 30, 45, 50, 65, 80, 95, 100)

# Scale to [0,1] range
scaled_data <- scale_values(original_data)

kable(data.frame(Original = original_data, Scaled = scaled_data), align = "c")
# Check the range
cat("Scaled range:", range(scaled_data), "\n")
```
